From Precipitation to Prediction: Integrating ConvLSTM Models for Comprehensive River Runoff Forecasting

### 2024-05-27 paper/paper.tex
went through text

\section{Introduction}\label{introduction}

@@

In this work, we demonstrate that ConvLSTM networks are a reliable
method for predicting multiple rivers simultaneously, using only
atmospheric forcing, even in the absence of a fully functional
hydrological model with a complex parameterization. We use the Baltic
Sea catchment as an example to illustrate our approach. Although the
methodology we propose is universally applicable across various
geographic regions, the Baltic Sea presents a challenge due to its
unique hydrological characteristics, being nearly decoupled from the
open ocean (see Figure). As a consequence, the salinity of the Baltic
Sea is driven to a large part by freshwater supply from rivers.
Freshwater enters the Baltic Sea through river runoff or positive net
precipitation (precipitation minus evaporation) over the sea surface.
The net precipitation accounts for 11 \% and the river input for 89 \%
of the total freshwater input (Meier \& Döscher, 2002). Modelling the
Baltic Sea is therefore to a large part the result of the quality of the
river input, that is used for the simulation. This makes the accurate
modeling of river runoff especially critical for simulations pertaining
to the Baltic Sea.

@@

\section{Implemented model
architecture}\label{implemented-model-architecture}

\subsection{The main idea}\label{sec-main_idea}

We assume that the runoff for a specific point in time for all \(N_r\)
considered rivers, collected in the vector
\(\vec{R}^t \in \mathbb{R}^{N_r}\), can be accurately approximated by a
functional \(\vec{M}(\{X^t_k[x,y,\tau]\})\) of \(k=1,\ldots N_k\)
atmospheric fields \(X^t_k[x,y,\tau]\) that are known for the past
\(\tau=1,\ldots N_\tau\) time instances, i.e.

@@

The fields are evaluated on a spatial domain \(x=1,\ldots N_x\) and
\(y=1,\ldots N_y\) that is sufficiently large to capture all significant
local and non-local contributions of atmospheric fields to the river
runoff.

Usually such a mapping is realized in a suitable hydrological model that
simulates all relevant physical processes that transform variables like,
e.g., precipitation and evaproration to the river runoff, relying on
expertise for tuning all appearing parameters to reasonable values.

As an alternative, we propose that this functional can be adequately
represented by a combination of a ConvLSTM with a subsequent FC network.
That way, there is no need to have knowledge of involved processes and
how they can be modeled, instead, these features can be ``lerned'' by
the network in an automated manner. Our poroposed network architecture
is visualized in Figure~\ref{fig-ConvLSTM_FC} and described in detail in
the following. In order to provide an overview, we will go through the
main ingredients of this architecture one-by-one.

@@

\centering{

\includegraphics{./ConvLSTM_FC.png}

@@

\end{figure}%

\subsection{ConvLSTM network}\label{convlstm-network}

\subsubsection{The LSTM approach}\label{sec-LSTM}

Before turning directly to the convolutional LSTM, the simpler
architecture of the plain LSTM is examined and should serve as a role
model for the later consideration of the full ConvLSTM.

The Long Short-Term Memory (LSTM), a specialized form of Recurrent
Neural Networks (RNNs), is specifically tailored for modeling temporal
sequences
\(\vec{X}^t[1], \ldots \vec{X}^t[\tau],\ldots \vec{X}^t[N_\tau]\) of
\(N_\tau\) input quantities
\(\vec{X}^t[\tau] = (X^t_k[\tau]) \in \mathbb{R}^{N_k}\). The sequence
is taken from a dataset given in form of a time series \(\{\vec{X}^t\}\)
and the endpoint should coincide with the specific element in the time
series connected to time \(t\),
i.e.~\(\vec{X}^t[N_\tau] \equiv \vec{X}^t\), see
Figure~\ref{fig-ConvLSTM_FC}. The \(N_k\) is the number of used input
``channels'' and can, for example, represent different measurable
quantities. Its unique design allows it to adeptly handle long-range
dependencies, setting it apart from traditional RNNs in terms of
accuracy (see Figure~\ref{fig-lstm}).

@@

\centering{

\includegraphics{./LSTM.png}

@@

In particular, the forget gate \(\vec{F}^t[\tau]\) defines the portion
of the previous (long-term memory) cell state \(\vec{C}^t[\tau-1]\) that
should be kept, see dashed cyan box therin\hspace{0pt}. The input gate
\(\vec{I}^t[\tau]\) controls the contribution of the current input used
to update the long-term memory, \(\vec{C}^t[\tau]\) (magenta and yellow
boxes). The output gate, \(\vec{O}^t[\tau]\), then determines how much
of this updated long-term memory contributes to the new (short-term
memory) hidden state, \(\vec{H}^t[\tau]\) (black dashed box).

@@

where \(g=i,f,o,c\) stands for the input, forget, output and cell-state
gate, respectively and Eistein's summation convention is employed,
i.e.~indices that appear twice are summed over. The calligraphic symbols
\(\mathcal{M}^{g}_{hk}, \mathcal{N}^{g}_{hh'}\) and
\(\mathcal{B}^g_{h}\) are the free parameters of the network that are
optimized for the given problem during the training, which is at the
heart of the any machine learning approach. The matrix
\(\pmb{\mathcal{M}}^{g} = (\mathcal{M}^{g}_{hk}) \in \mathbb{R}^{N_h \times N_k}\)
can be interpreted as a Markovian-like contribution of the current input
\(\vec{X}^t[\tau]\) to the gates, whereas the
\(\pmb{\mathcal{N}}^{g} = (\mathcal{N}^{g}_{hh'}) \in \mathbb{R}^{N_h \times N_h}\)
scales a non-Markovian part determined by the hidden state of the last
sequence point \(\tau-1\). The vector
\(\vec{\mathcal{B}}^g = (\mathcal{B}^g_{h}) \in \mathbb{R}^{N_h}\) is a
learnable bias. It should be stressed that these parameters do neither
depend on \(t\) nor on \(\tau\) and are thus optimized once for the
complete dataset \(\{\vec{X}^t\}\).

@@

For the sake of brevity, we can write the mapping more compactly in
matrix-vector form as
\begin{equation}\phantomsection\label{eq-LSTM-gates}{
\begin{aligned}
\vec{g}^t[\tau] = \pmb{\mathcal{M}}^{g}\vec{X}^t[\tau] + \pmb{\mathcal{N}}^{g}  \vec{H}^t[ \tau-1]+ \vec{\mathcal{B}}^g \,
\end{aligned}
}\end{equation}

@@

\begin{equation}\phantomsection\label{eq-LSTM}{
\begin{aligned}
\vec{I}^t[\tau] &= \sigma(\vec{i}^t[\tau]) \\
\vec{F}^t[\tau] &= \sigma(\vec{f}^t[\tau]) \\
\vec{O}^t[\tau] &= \sigma(\vec{o}^t[\tau]) \\
\vec{C}^t[\tau] &= \vec{F}^t[\tau] \circ \vec{C}^t[\tau-1]  + \vec{I}^t[\tau] \circ \tanh(\vec{c}^t[\tau]) \\
\vec{H}^t[\tau]  &= \vec{O}^t[\tau] \circ \tanh(\vec{C}^t[\tau]) \ ,
\end{aligned}
}\end{equation}

where \(\sigma\) denotes the logisitc sigmoid function, \(\tanh\) is the
hyperbolic tangent and the \(\circ\) stands for the Hadamard product
(all applied in an element-wise fashion to the vectors). In the last two
equations the role of the input, forget and output gates as described
above becomes apparent.

@@

\begin{equation}\phantomsection\label{eq-L_nested}{
\left(\vec{H}^t[N_\tau], \vec{C}^t[N_\tau] \right) = L \left( \vec{X}^t[N_\tau], L \left( \vec{X}^t[N_\tau-1], \ldots L \left( \vec{X}^t[1], (\vec{H}^t[0], \vec{C}^t[0]) \right) \ldots \right) \right) \ ,
}\end{equation}

@@

\subsubsection{Combining LSTM with spatial
convolution}\label{combining-lstm-with-spatial-convolution}

Although the plain LSTM has high performance in handling temporal
sequences of point-like quantities it is not designed to recognize
spatial features in seuqences of, e.g., two-dimensional maps as
atmospheric-ocean interface fields. To adress this limitation we employ
a ConvLSTM architecture as described in the following.

In this kind of network the elements of the input sequence are given as
spatially varying fields
\(\vec{X}^t[\tau] = (X^t_k[x,y,\tau]) \in \mathbb{R}^{N_k \times (N_x \times N_y)}\),
where \(x\in[1, N_x]\) and \(y \in [1, N_y]\) run over the horizontal
and vertical dimensions of the map, respectively. In order to enable the
``learning'' of spatial patterns, the free parameters of the network are
replaced by two-dimenmsional convolutional kernels
\(\pmb{\mathcal{M}}^{g} = (\mathcal{M}^{g}_{hk}[\xi, \eta]) \in \mathbb{R}^{(N_h \times N_k)\times (N_\xi \times N_\eta)}\)
and
\(\pmb{\mathcal{N}}^{g} = (\mathcal{N}^{g}_{hh'}[\xi, \eta]) \in \mathbb{R}^{(N_h \times N_h)\times (N_\xi \times N_\eta)}\).
The width and the height of the kernels are given by \(N_\xi\) and
\(N_\eta\), respectively and
\(\xi\in [-(N_\xi-1)/2,(N_\xi-1)/2], \eta\in [-(N_\eta-1)/2,(N_\eta-1)/2]\),
where, without loss of generality, we assume odd numbers for the kernel
sizes.

@@

It becomes immediately apparent that in case of the ConvLSTM, the gate
and state vectors must become vector fields
(\(\in \mathbb{R}^{N_h \times (N_x \times N_y)}\)) as well. We can write
this mapping in the same fashion as Equation~\ref{eq-LSTM-gates} but
with replacing the normal matrix-vector multiplication by a convolution
(denoted with \(\ast\)), i.e.\\
\[
\begin{aligned}
\vec{g}^t[\tau] = \pmb{\mathcal{M}}^{g} \ast \vec{X}^t[\tau] + \pmb{\mathcal{N}}^{g} \ast  \vec{H}^t[ \tau-1]+ \vec{\mathcal{B}}^g \,
\end{aligned}
\]

The susebsequent processing of the \(\vec{g}^t[\tau]\) remains
symbolically the same as presented in Equation~\ref{eq-LSTM} but with
all appearing quantities now meaning vector fields instead of simple
vectors.

@@

\subsection{Fully connected layer}\label{sec-FC_layer}

@@

\section{Technical details}\label{sec-technical_details}

\subsection{Runoff data used for
training}\label{runoff-data-used-for-training}

The runoff data covering the period 1979 to 2011 is based on an E-HYPE
hindcast simulation that was forced by a regional downscaling of
ERA-Interim (Dee et al., 2011) with RCA3 ({``The rossby centre regional
climate model RCA3,''} n.d.) and implemented into NEMO-Nordic (Hordoir
et al., 2019) as a mass flux. For the periods before (1961 to 1978) and
after (2012 to 2018) additional spatial temporal corrections have been
applied to the runoff data, and have therefore been ignored. For more
information see Gröger et al. (2022) and references therein.

\subsection{Atmospheric Forcing}\label{atmospheric-forcing}

@@

\subsection{Ocean Model}\label{ocean-model}

@@

\subsection{Neural network hyper
parameters}\label{neural-network-hyper-parameters}

@@

\centering{

\includegraphics{ConvLSTM.png}

@@

\begin{longtable}[]{@{}ll@{}}
\caption{Hyperparameters}\label{tbl-letters}\tabularnewline
\toprule\noalign{}
Parameter name & Parameter size \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Parameter name & Parameter size \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Channel size & 4 \\
Num. hidden layer & 9 \\
Num. timesteps & 30 \\
Conv. Kernelsize & (7,7) \\
Num. ConvLSTM layers & 1 \\
Batch size & 50 \\
Learning Rate & 1e-3 with CosineAnnealing \\
\end{longtable}

@@

The choice of atmospheric fields was based on the implemented river
runoff calculation in the atmospheric model COSMO-CLM which uses these
four fields to calculate an river runoff estimate.

\section{Results}\label{results}

The model was trained with daily data for the period 1979 to 2011, as
this period represents the only period of E-HYPE without further bias
correction applied to the runoff to match observations. The data was
divided into randomly chosen splits of 80\% training data, 10\%
validation data to evaluate the performance of the model during
training, and 10\% training data which is finally used to evaluate the
performance of the model after training. The model was trained for 400
epochs and the model weights with the lowest mean squared error during
training have been stored.

The accuracy of the model is displayed in Figure
\textbf{?@fig-statistical-evaluationNN}. As mention above for
evaluation, the test dataset was utilized. The left panel panel (Figure
\textbf{?@fig-statistical-evaluationNN} a) ) illustrates the relative
prediction error in relation to the original E-HYPE data, indicating
that, on daily timescales, the model can predict river runoff with an
accuracy of \(\pm\) 5\%. The overall correlation is 0.997 with the
resulting error metrics yielding a RMSE of 323.99 \(m^3\)/s and MAE of
249.51 \(m^3\)/s. While the model's performance is satisfactory, the
discrepancies between the actual values and the predictions could partly
be attributed to the use of a different atmospheric dataset than the one
originally used to drive the E-HYPE model. However, by applying a
rolling mean with a 5-day window, the prediction error is reduced to
less than 1\%, which is acceptable for the purposes of climate modeling.
Lastly, the right panel demonstrates that the distribution of residuals
follows a Gaussian shape, suggesting that our model does not exhibit
bias by systematically over or underestimating river runoff values.

@@

The final evaluation of the ConvLSTM model concentrates on the spatial
accuracy of river runoff predictions. Figure xxx a) shows the vertically
averaged salinity for the period 1981 to 2011 in the reference
simulation. It highlights the strong horizontal gradients and complex
topographic features in the Baltic Sea, as evidenced by salinity
variations in deeper waters, captured by the vertical integration.
Figure 4b compares these results by showing the percentage difference in
vertically averaged salinity between the ConvLSTM simulation and the
reference simulation. Overall, the differences remain below 1\%, except
near a river mouth in the Gulf of Riga (22-24°E, 56.5-58.5°N for
orientation), where the difference is approximately 1\%.

@@

\section{Discussion}\label{discussion}

@@

In summary, we showed that the ConvLSTM model demonstrated robust
performance in forecasting river runoff across 97 rivers entering the
Baltic Sea. Trained on data from 1979 to 2011, the model achieved an
impressive daily prediction accuracy of ±5\%. This capability to
generate accurate and detailed simulations enables to examine the
potential impacts of different climate scenarios. Such precision in
forecasting and scenario testing is crucial for crafting effective water
resource management strategies and adapting to the changing climate.

@@

\section{Acknowledgments}\label{acknowledgments}

@@

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\vspace{1em}

\bibitem[\citeproctext]{ref-ashrafi2017fully}
Ashrafi, M., Chua, L. H. C., Quek, C., \& Qin, X. (2017). A fully-online
neuro-fuzzy model for flow forecasting in basins with limited data.
\emph{Journal of Hydrology}, \emph{545}, 424--435.

\bibitem[\citeproctext]{ref-dee2011}
Dee, D. P., Uppala, S. M., Simmons, A. J., Berrisford, P., Poli, P.,
Kobayashi, S., et al. (2011). The ERA-Interim reanalysis: configuration
and performance of the data assimilation system. \emph{Quarterly Journal
of the Royal Meteorological Society}, \emph{137}(656), 553--597.
\url{https://doi.org/10.1002/qj.828}

\bibitem[\citeproctext]{ref-fang2022application}
Fang, L., \& Shao, D. (2022). Application of long short-term memory
(LSTM) on the prediction of rainfall-runoff in karst area.
\emph{Frontiers in Physics}, \emph{9}, 790687.

\bibitem[\citeproctext]{ref-fangExaminingApplicabilityDifferent2019}
Fang, W., Huang, S., Ren, K., Huang, Q., Huang, G., Cheng, G., \& Li, K.
(2019). Examining the applicability of different sampling techniques in
the development of decomposition-based streamflow forecasting models.
\emph{Journal of Hydrology}, \emph{568}, 534--550.
\url{https://doi.org/10.1016/j.jhydrol.2018.11.020}

\bibitem[\citeproctext]{ref-gruxf6ger2022}
Gröger, M., Placke, M., Meier, M., Börgel, F., Brunnabend, S.-E.,
Dutheil, C., et al. (2022). \emph{The Baltic Sea model inter-comparison
project BMIP {\textendash} a platform for model development, evaluation,
and uncertainty assessment}. Retrieved from
\url{https://gmd.copernicus.org/preprints/gmd-2022-160/}

\bibitem[\citeproctext]{ref-ha2021}
Ha, S., Liu, D., \& Mu, L. (2021). Prediction of Yangtze River
streamflow based on deep learning neural network with El
Niño{\textendash}Southern Oscillation. \emph{Scientific Reports},
\emph{11}(1), 11738. \url{https://doi.org/10.1038/s41598-021-90964-3}

\bibitem[\citeproctext]{ref-hagemannHighResolutionDischarge2020}
Hagemann, S., Stacke, T., \& Ho-Hagemann, H. T. M. (2020). High
{Resolution Discharge Simulations Over Europe} and the {Baltic Sea
Catchment}. \emph{Frontiers in Earth Science}, \emph{8}.
\url{https://doi.org/10.3389/feart.2020.00012}

\bibitem[\citeproctext]{ref-hochreiter1997long}
Hochreiter, S., \& Schmidhuber, J. (1997). Long short-term memory.
\emph{Neural Computation}, \emph{9}(8), 1735--1780.

>>>>

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, interior hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, frame hidden, boxrule=0pt, sharp corners]}{\end{tcolorbox}}\fi

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

@@

In this work, we demonstrate that ConvLSTM networks are a reliable
method for predicting multiple rivers simultaneously, using only
atmospheric forcing, even in the absence of a fully functional
hydrological model with a complex parameterization. We use the Baltic
Sea catchment as an example to illustrate our approach. Although the
methodology we propose is universally applicable across various
geographic regions, the Baltic Sea presents a challenge due to its
unique hydrological characteristics, being nearly decoupled from the
open ocean (see Figure). As a consequence, the salinity of the Baltic
Sea is driven to a large part by freshwater supply from rivers.
Freshwater enters the Baltic Sea through river runoff or positive net
precipitation (precipitation minus evaporation) over the sea surface.
The net precipitation accounts for 11 \(\%\) and the river input for 89
\(\%\) of the total freshwater input (Meier \& Döscher, 2002). Modelling
the Baltic Sea is therefore to a large part the result of the quality of
the river input, that is used for the simulation. This makes the
accurate modeling of river runoff especially critical for simulations
pertaining to the Baltic Sea.

@@

\hypertarget{implemented-model-architecture}{%
\section{Implemented model
architecture}\label{implemented-model-architecture}}

\hypertarget{sec-main_idea}{%
\subsection{The main idea}\label{sec-main_idea}}

We assume that the runoff at a specific point in time for all \(N_r\)
considered rivers, collected in the vector
\(\vec{R}^t \in \mathbb{R}^{N_r}\), can be accurately approximated by a
functional \(\vec{M}(\{X^t_k[x,y,\tau]\})\) of \(k=1,\ldots N_k\)
atmospheric fields \(X^t_k[x,y,\tau]\) which are known for the past
\(\tau=1,\ldots N_\tau\) time instances. This relationship is expressed
as:

@@

The atmospheric fields are evaluated over a spatial domain
\(x=1,\ldots N_x\) and \(y=1,\ldots N_y\) which is sufficiently large to
capture all significant local and non-local contributions of the
atmospheric fields to the river runoff.

Typically, such a mapping is realized using a hydrological model that
simulates all relevant physical processes, transforming variables like
precipitation and evaporation into river runoff. This process relies
heavily on domain knowledge to tune all parameters to reasonable values.

As an alternative, we propose that this functional can be adequately
represented by a combination of a ConvLSTM with a subsequent FC network.
This approach eliminates the need for detailed knowledge of the involved
processes and their modeling. Instead, these features can be ``learned''
by the network in an automated manner. Our proposed network architecture
is visualized in Figure~\ref{fig-ConvLSTM_FC} and described in detail in
the following sections. To provide an overview, we will discuss the main
components of this architecture one-by-one.

@@

{\centering \includegraphics{./ConvLSTM_FC.png}

@@

\end{figure}

\hypertarget{convlstm-network}{%
\subsection{ConvLSTM network}\label{convlstm-network}}

\hypertarget{sec-LSTM}{%
\subsubsection{The LSTM approach}\label{sec-LSTM}}

Before turning directly to the convolutional LSTM, the simpler
architecture of the plain LSTM is examined which serves as a foundation
for understanding the more complex ConvLSTM.

The Long Short-Term Memory (LSTM), a specialized form of Recurrent
Neural Networks (RNNs), is specifically designed to model temporal
sequences
\(\vec{X}^t[1], \ldots \vec{X}^t[\tau],\ldots \vec{X}^t[N_\tau]\) of
\(N_\tau\) input quantities
\(\vec{X}^t[\tau] = (X^t_k[\tau]) \in \mathbb{R}^{N_k}\). This sequence
is taken from a dataset given in form of a time series \(\{\vec{X}^t\}\)
with the endpoint coinciding with the specific element in the time
series connected to time \(t\),
i.e.~\(\vec{X}^t[N_\tau] \equiv \vec{X}^t\), see
Figure~\ref{fig-ConvLSTM_FC}. Here, \(N_k\) represents the number of
input ``channels,'' which can correspond to different measurable
quantities. The LSTM's unique design allows it to adeptly handle
long-range dependencies, setting it apart from traditional RNNs in terms
of accuracy (see Figure~\ref{fig-lstm}).

@@

{\centering \includegraphics{./LSTM.png}

@@

In particular, the forget gate \(\vec{F}^t[\tau]\) defines the portion
of the previous (long-term memory) cell state \(\vec{C}^t[\tau-1]\) that
should be kept, see dashed cyan box therein\hspace{0pt}. The input gate
\(\vec{I}^t[\tau]\) controls the contribution of the current input used
to update the long-term memory, \(\vec{C}^t[\tau]\) (magenta and yellow
boxes). The output gate, \(\vec{O}^t[\tau]\), then determines how much
of this updated long-term memory contributes to the new (short-term
memory) hidden state, \(\vec{H}^t[\tau]\) (black dashed box).

@@

where \(g=i,f,o,c\) stands for the input, forget, output and cell-state
gate, respectively and Einstein's summation convention is employed,
i.e.~indices that appear twice are summed over. The calligraphic symbols
\(\mathcal{M}^{g}_{hk}, \mathcal{N}^{g}_{hh'}\) and
\(\mathcal{B}^g_{h}\) are the free parameters of the network that are
optimized for the given problem during the training, which is at the
heart of any machine learning approach. The matrix
\(\pmb{\mathcal{M}}^{g} = (\mathcal{M}^{g}_{hk}) \in \mathbb{R}^{N_h \times N_k}\)
can be interpreted as a Markovian-like contribution of the current input
\(\vec{X}^t[\tau]\) to the gates, whereas the
\(\pmb{\mathcal{N}}^{g} = (\mathcal{N}^{g}_{hh'}) \in \mathbb{R}^{N_h \times N_h}\)
scales a non-Markovian part determined by the hidden state of the last
sequence point \(\tau-1\). The vector
\(\vec{\mathcal{B}}^g = (\mathcal{B}^g_{h}) \in \mathbb{R}^{N_h}\) is a
learnable bias. It should be stressed that these parameters do neither
depend on \(t\) nor on \(\tau\) and are thus optimized once for the
complete dataset \(\{\vec{X}^t\}\).

@@

For the sake of brevity, we can write the mapping more compactly in
matrix-vector form as
\begin{equation}\protect\hypertarget{eq-LSTM-gates}{}{
\begin{aligned}
\vec{g}^t[\tau] = \pmb{\mathcal{M}}^{g}\vec{X}^t[\tau] + \pmb{\mathcal{N}}^{g}  \vec{H}^t[ \tau-1]+ \vec{\mathcal{B}}^g \,
\end{aligned}
}\label{eq-LSTM-gates}\end{equation}

@@

\begin{equation}\protect\hypertarget{eq-LSTM}{}{
\begin{aligned}
\vec{I}^t[\tau] &= \sigma(\vec{i}^t[\tau]) \\
\vec{F}^t[\tau] &= \sigma(\vec{f}^t[\tau]) \\
\vec{O}^t[\tau] &= \sigma(\vec{o}^t[\tau]) \\
\vec{C}^t[\tau] &= \vec{F}^t[\tau] \circ \vec{C}^t[\tau-1]  + \vec{I}^t[\tau] \circ \tanh(\vec{c}^t[\tau]) \\
\vec{H}^t[\tau]  &= \vec{O}^t[\tau] \circ \tanh(\vec{C}^t[\tau]) \ ,
\end{aligned}
}\label{eq-LSTM}\end{equation}

where \(\sigma\) denotes the logistic sigmoid function, \(\tanh\) is the
hyperbolic tangent and the \(\circ\) stands for the Hadamard product
(all applied in an element-wise fashion to the vectors). In the last two
equations the role of the input, forget and output gates as described
above becomes apparent.

@@

\begin{equation}\protect\hypertarget{eq-L_nested}{}{
\left(\vec{H}^t[N_\tau], \vec{C}^t[N_\tau] \right) = L \left( \vec{X}^t[N_\tau], L \left( \vec{X}^t[N_\tau-1], \ldots L \left( \vec{X}^t[1], (\vec{H}^t[0], \vec{C}^t[0]) \right) \ldots \right) \right) \ ,
}\label{eq-L_nested}\end{equation}

@@

\hypertarget{combining-lstm-with-spatial-convolution}{%
\subsubsection{Combining LSTM with spatial
convolution}\label{combining-lstm-with-spatial-convolution}}

Although the plain LSTM has high performance in handling temporal
sequences of point-like quantities it is not designed to recognize
spatial features in sequences of, e.g., two-dimensional maps as
atmospheric-ocean interface fields. To address this limitation we employ
a ConvLSTM architecture as described in the following.

In this kind of network the elements of the input sequence are given as
spatially varying fields
\(\vec{X}^t[\tau] = (X^t_k[x,y,\tau]) \in \mathbb{R}^{N_k \times (N_x \times N_y)}\),
where \(x\in[1, N_x]\) and \(y \in [1, N_y]\) run over the horizontal
and vertical dimensions of the map, respectively. In order to enable the
``learning'' of spatial patterns, the free parameters of the network are
replaced by two-dimensional convolution kernels
\(\pmb{\mathcal{M}}^{g} = (\mathcal{M}^{g}_{hk}[\xi, \eta]) \in \mathbb{R}^{(N_h \times N_k)\times (N_\xi \times N_\eta)}\)
and
\(\pmb{\mathcal{N}}^{g} = (\mathcal{N}^{g}_{hh'}[\xi, \eta]) \in \mathbb{R}^{(N_h \times N_h)\times (N_\xi \times N_\eta)}\).
The width and the height of the kernels are given by \(N_\xi\) and
\(N_\eta\), respectively and
\(\xi\in [-(N_\xi-1)/2,(N_\xi-1)/2], \eta\in [-(N_\eta-1)/2,(N_\eta-1)/2]\),
where, without loss of generality, we assume odd numbers for the kernel
sizes.

@@

It becomes immediately apparent that in case of the ConvLSTM, the gate
and state vectors must become vector fields
(\(\in \mathbb{R}^{N_h \times (N_x \times N_y)}\)) as well. We can write
this mapping in the same way as Equation~\ref{eq-LSTM-gates} but with
replacing the normal matrix-vector multiplication by a convolution
(denoted with \(\ast\)), i.e.\\
\[
\begin{aligned}
\vec{g}^t[\tau] = \pmb{\mathcal{M}}^{g} \ast \vec{X}^t[\tau] + \pmb{\mathcal{N}}^{g} \ast  \vec{H}^t[ \tau-1]+ \vec{\mathcal{B}}^g \,
\end{aligned}
\]

The subsequent processing of the \(\vec{g}^t[\tau]\) remains
symbolically the same as presented in Equation~\ref{eq-LSTM} but with
all appearing quantities now meaning vector fields instead of simple
vectors.

@@

\hypertarget{sec-FC_layer}{%
\subsection{Fully connected layer}\label{sec-FC_layer}}

@@

\hypertarget{sec-technical_details}{%
\section{Technical details}\label{sec-technical_details}}

\hypertarget{runoff-data-used-for-training}{%
\subsection{Runoff data used for
training}\label{runoff-data-used-for-training}}

The runoff data covering the period 1979 to 2011 is based on an E-HYPE
hindcast simulation that was forced by a regional downscaling of
ERA-Interim (Dee et al., 2011) with RCA3 ({``The rossby centre regional
climate model RCA3,''} n.d.) and implemented into NEMO-Nordic (Hordoir
et al., 2019) as a mass flux. For the periods before (1961 to 1978) and
after (2012 to 2018) additional spatial temporal corrections have been
applied to the runoff data, and have therefore been ignored. The quality
of the runoff was extensively evaluated. For more information see Gröger
et al. (2022) and references therein.

\hypertarget{atmospheric-forcing}{%
\subsection{Atmospheric Forcing}\label{atmospheric-forcing}}

@@

\hypertarget{ocean-model}{%
\subsection{Ocean Model}\label{ocean-model}}

@@

\hypertarget{neural-network-hyper-parameters}{%
\subsection{Neural network hyper
parameters}\label{neural-network-hyper-parameters}}

@@

{\centering \includegraphics{ConvLSTM.png}

@@

\hypertarget{tbl-letters}{}
\begin{longtable}[]{@{}ll@{}}
\caption{\label{tbl-letters}Hyperparameters}\tabularnewline
\toprule()
Parameter name & Parameter size \\
\midrule()
\endfirsthead
\toprule()
Parameter name & Parameter size \\
\midrule()
\endhead
Channel size & 4 \\
Num. hidden layer & 9 \\
Num. timesteps & 30 \\
Conv. Kernelsize & (7,7) \\
Num. ConvLSTM layers & 1 \\
Batch size & 50 \\
Learning Rate & 1e-3 with CosineAnnealing \\
\bottomrule()
\end{longtable}

@@

The choice of atmospheric fields was based on the implemented river
runoff calculation in the atmospheric model COSMO-CLM which uses these
atmospheric fields to calculate an river runoff estimate.

\hypertarget{results}{%
\section{Results}\label{results}}

The model was trained with daily data for the period 1979 to 2011, as
this period represents the only period of E-HYPE without further bias
correction applied to the runoff to match observations. The data was
divided into randomly chosen splits of 80\(\%\) training data, 10\(\%\)
validation data to evaluate the performance of the model during
training, and 10\(\%\) training data which is finally used to evaluate
the performance of the model after training. The model was trained for
400 epochs and the model weights with the lowest mean squared error
during training have been stored.

The accuracy of the model is displayed in Figure
\textbf{?@fig-statistical-evaluationNN}. As mention above for
evaluation, the test dataset was utilized. The left panel panel (Figure
\textbf{?@fig-statistical-evaluationNN} a) ) illustrates the relative
prediction error in relation to the original E-HYPE data, indicating
that, on daily timescales, the model can predict river runoff with an
accuracy of \(\pm\) 5\(\%\). The overall correlation is 0.997 with the
resulting error metrics yielding a RMSE of 323.99 \(m^3\)/s and MAE of
249.51 \(m^3\)/s. While the model's performance is satisfactory, the
discrepancies between the actual values and the predictions could partly
be attributed to the use of a different atmospheric dataset than the one
originally used to drive the E-HYPE model. However, by applying a
rolling mean with a 5-day window, the prediction error is reduced to
less than 1\(\%\), which is acceptable for the purposes of climate
modeling. Lastly, the right panel demonstrates that the distribution of
residuals follows a Gaussian shape, suggesting that our model does not
exhibit bias by systematically over or underestimating river runoff
values.

@@

The final evaluation of the ConvLSTM model concentrates on the spatial
accuracy of river runoff predictions. Figure xxx a) shows the vertically
averaged salinity for the period 1981 to 2011 in the reference
simulation. It highlights the strong horizontal gradients and complex
topographic features in the Baltic Sea, as evidenced by salinity
variations in deeper waters, captured by the vertical integration.
Figure 4b compares these results by showing the percentage difference in
vertically averaged salinity between the ConvLSTM simulation and the
reference simulation. Overall, the differences remain below 1\(\%\),
except near a river mouth in the Gulf of Riga (22-24°E, 56.5-58.5°N for
orientation), where the difference is approximately 1\(\%\).

@@

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

@@

In summary, we showed that the ConvLSTM model demonstrated robust
performance in forecasting river runoff across 97 rivers entering the
Baltic Sea. Trained on data from 1979 to 2011, the model achieved an
impressive daily prediction accuracy of ±5\(\%\). This capability to
generate accurate and detailed simulations enables to examine the
potential impacts of different climate scenarios. Such precision in
forecasting and scenario testing is crucial for crafting effective water
resource management strategies and adapting to the changing climate.

@@

\hypertarget{acknowledgments}{%
\section{Acknowledgments}\label{acknowledgments}}

@@

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\vspace{1em}

\leavevmode\vadjust pre{\hypertarget{ref-ashrafi2017fully}{}}%
Ashrafi, M., Chua, L. H. C., Quek, C., \& Qin, X. (2017). A fully-online
neuro-fuzzy model for flow forecasting in basins with limited data.
\emph{Journal of Hydrology}, \emph{545}, 424--435.

\leavevmode\vadjust pre{\hypertarget{ref-dee2011}{}}%
Dee, D. P., Uppala, S. M., Simmons, A. J., Berrisford, P., Poli, P.,
Kobayashi, S., et al. (2011). The ERA-Interim reanalysis: configuration
and performance of the data assimilation system. \emph{Quarterly Journal
of the Royal Meteorological Society}, \emph{137}(656), 553--597.
\url{https://doi.org/10.1002/qj.828}

\leavevmode\vadjust pre{\hypertarget{ref-fang2022application}{}}%
Fang, L., \& Shao, D. (2022). Application of long short-term memory
(LSTM) on the prediction of rainfall-runoff in karst area.
\emph{Frontiers in Physics}, \emph{9}, 790687.

\leavevmode\vadjust pre{\hypertarget{ref-fangExaminingApplicabilityDifferent2019}{}}%
Fang, W., Huang, S., Ren, K., Huang, Q., Huang, G., Cheng, G., \& Li, K.
(2019). Examining the applicability of different sampling techniques in
the development of decomposition-based streamflow forecasting models.
\emph{Journal of Hydrology}, \emph{568}, 534--550.
\url{https://doi.org/10.1016/j.jhydrol.2018.11.020}

\leavevmode\vadjust pre{\hypertarget{ref-gruxf6ger2022}{}}%
Gröger, M., Placke, M., Meier, M., Börgel, F., Brunnabend, S.-E.,
Dutheil, C., et al. (2022). \emph{The Baltic Sea model inter-comparison
project BMIP {\textendash} a platform for model development, evaluation,
and uncertainty assessment}. Retrieved from
\url{https://gmd.copernicus.org/preprints/gmd-2022-160/}

\leavevmode\vadjust pre{\hypertarget{ref-ha2021}{}}%
Ha, S., Liu, D., \& Mu, L. (2021). Prediction of Yangtze River
streamflow based on deep learning neural network with El
Niño{\textendash}Southern Oscillation. \emph{Scientific Reports},
\emph{11}(1), 11738. \url{https://doi.org/10.1038/s41598-021-90964-3}

\leavevmode\vadjust pre{\hypertarget{ref-hagemannHighResolutionDischarge2020}{}}%
Hagemann, S., Stacke, T., \& Ho-Hagemann, H. T. M. (2020). High
{Resolution Discharge Simulations Over Europe} and the {Baltic Sea
Catchment}. \emph{Frontiers in Earth Science}, \emph{8}.
\url{https://doi.org/10.3389/feart.2020.00012}
